
---
title: "Interpreting Gas Production Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting Gas Production Models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Fitting a model is only the first step in the analysis
of rumen gas production data.

Researchers must also interpret:

- Model parameters
- Biological meaning
- Goodness-of-fit statistics
- Competing model performance

This vignette summarizes the most common
interpretations used in rumen gas production studies.

```{r}
library(rumenGP)
```

# Understanding Common Parameters

Although different models use different equations,
many share similar biological concepts.

---

# Asymptotic Gas Production

Common parameter names:

```text
A
VF
Vf
V1F
V2F
```

These parameters represent the maximum gas
production that the model predicts after
long incubation times.

Example:

```text
A = 120 mL
```

Interpretation:

```text
The model predicts approximately
120 mL of gas at fermentation completion.
```

Higher values generally indicate:

- Greater fermentable substrate availability
- Increased fermentation potential

However, interpretation should always be made
within the context of the substrate being studied.

---

# Fermentation Rate

Common parameter names:

```text
k
k1
k2
mu
r
```

These parameters describe how rapidly gas
production approaches the asymptote.

Example:

```text
Treatment A
k = 0.08

Treatment B
k = 0.04
```

Interpretation:

```text
Treatment A ferments more rapidly
than Treatment B.
```

Higher rates generally suggest:

- Faster microbial degradation
- Greater substrate accessibility
- More rapid attainment of asymptotic gas production

In Burr XII and Inverse Paralogistic models,
the parameter:

```text
r
```

serves a similar role.

---

# Lag Time

Common parameter name:

```text
lambda
```

or

\[
\lambda
\]

Lag time represents the delay before
substantial fermentation begins.

Example:

```text
lambda = 2 h
```

Interpretation:

```text
Approximately two hours are required
before active fermentation starts.
```

Large lag values often occur with:

- Fibrous substrates
- Physically protected nutrients
- Slowly colonized feeds

---

# Half-Time Parameters

Common parameter names:

```text
b
K
```

Used in:

- Groot
- Generalized Michaelis-Menten

These parameters determine the time required
to achieve approximately half of the asymptotic
gas production.

Example:

```text
K = 12 h
```

Interpretation:

```text
Approximately 50% of total gas production
is achieved after 12 hours.
```

Smaller values indicate faster fermentation.

---

# Shape Parameters

Common parameter names:

```text
a
c
d
k
m
p
```

Shape parameters modify the curvature of
the fermentation profile.

Interpretation:

```text
Shape parameters control how fermentation
accelerates and decelerates through time.
```

Unlike asymptotes or rates, shape parameters
often have no simple biological interpretation.

They are usually considered:

```text
Empirical flexibility parameters.
```

Examples:

### Groot

```text
k
```

controls curve steepness.

### Generalized Michaelis-Menten

```text
c
```

controls curve shape and steepness.

### Log-logistic

```text
a
```

controls curve shape and steepness.

### Burr XII

```text
a
p
```

jointly influence asymmetry,
curvature, and inflection behavior.

### Inverse Paralogistic

```text
a
```

controls the overall shape of the
fermentation profile.

---

# Interpreting Dual-Pool Models

Dual-pool models separate fermentation
into:

```text
Rapid fraction
Slow fraction
```

Parameters:

```text
V1F
V2F
k1
k2
```

---

## Rapid Fraction

```text
V1F
k1
```

Typically associated with:

- Soluble carbohydrates
- Readily fermentable compounds

---

## Slow Fraction

```text
V2F
k2
```

Typically associated with:

- Cell-wall components
- Structural carbohydrates
- Less accessible nutrients

Example:

```text
V1F = 30 mL

V2F = 90 mL
```

Interpretation:

```text
Most fermentation derives from
the slowly degradable fraction.
```

---

# Understanding Goodness-of-Fit Metrics

Model fit should never be evaluated
using a single statistic.

---

# R-Squared

\[
R^2
\]

Measures the proportion of observed
variation explained by the model.

Example:

```text
R² = 0.99
```

Interpretation:

```text
99% of variation is explained by
the fitted model.
```

### Important

A high R-squared does not guarantee
that the model is biologically meaningful
or scientifically preferable.

---

# RMSE

Root Mean Squared Error:

\[
RMSE
\]

Measures average prediction error.

Example:

```text
RMSE = 1.5 mL
```

Interpretation:

```text
Predictions differ from observations
by approximately 1.5 mL on average.
```

Smaller values are preferred.

---

# RSS

Residual Sum of Squares:

\[
RSS
\]

Represents total unexplained variation.

Smaller values indicate better fit.

---

# AIC

Akaike Information Criterion:

\[
AIC
\]

Balances:

```text
Fit quality
+
Model complexity
```

Smaller values are preferred.

AIC is especially useful when:

- Comparing non-nested models
- Comparing models with different numbers of parameters

---

# BIC

Bayesian Information Criterion:

\[
BIC
\]

Similar to AIC but applies a stronger
penalty for additional parameters.

Smaller values are preferred.

Because BIC penalizes complexity more heavily,
it often favors simpler models unless the
additional parameters substantially improve fit.

---

# Why Higher R² Does Not Always Mean a Better Model

Consider:

| Model | Parameters | R² | AIC |
|---------|---------|---------|---------|
| Groot | 3 | 0.9992 | 33 |
| Richards | 4 | 0.9994 | 35 |

The Richards model explains slightly more
variation.

However:

```text
Additional complexity
```

may not justify:

```text
Minimal improvement
```

AIC correctly penalizes the extra parameter.

Therefore:

```text
Higher R² alone should not determine
model selection.
```

---

# Understanding Model Equivalence

Several gas-production models are mathematically
equivalent despite using different parameter names.

### Groot

\[
V(t)
=
\frac{VF}
{
1+\left(\frac{b}{t}\right)^k
}
\]

### Generalized Michaelis-Menten

\[
V(t)
=
A
\frac{t^c}
{
t^c + K^c
}
\]

### Log-logistic

\[
V(t)
=
VF
\frac{(rt)^a}
{
1+(rt)^a
}
\]

Parameter correspondence:

| Groot | Generalized Michaelis-Menten | Log-logistic |
|---------|---------|---------|
| VF | A | VF |
| b | K | 1/r |
| k | c | a |

Therefore:

```text
Groot
=
Generalized Michaelis-Menten
=
Log-logistic
```

These formulations produce identical:

- Predicted values
- Residuals
- RSS
- RMSE
- R²
- AIC
- BIC

when equivalent parameter transformations are used.

Researchers may therefore choose the
parameterization most familiar within
their field.

---

# Interpreting Burr XII and Inverse Paralogistic Models

## Burr XII

Equation:

\[
V(t)
=
VF
\left[
1
-
\left(
1+(rt)^a
\right)^{-p}
\right]
\]

Key interpretation:

- VF determines asymptotic gas production
- r controls fermentation speed
- a and p jointly control shape and asymmetry

The model is highly flexible and can adapt
to many fermentation profiles.

However, this flexibility may increase
the risk of overfitting when datasets
are small.

---

## Inverse Paralogistic

Equation:

\[
V(t)
=
VF
\left[
1
+
(rt)^{-a}
\right]^{-a}
\]

Key interpretation:

- VF represents asymptotic gas production
- r influences production speed
- a controls curve shape and steepness

This model can describe diverse sigmoidal
profiles while retaining a relatively
simple parameter structure.

---

# Model Selection Strategy

Recommended workflow:

```text
1. Fit multiple models

2. Evaluate convergence

3. Compare RMSE

4. Compare AIC and BIC

5. Examine residual plots

6. Consider parameter plausibility

7. Consider biological interpretation

8. Select the most appropriate model
```

No single model should be considered
universally superior.

---

# Interpreting Failed Fits

Common reasons include:

```text
Poor starting values

Too many parameters

Insufficient observations

Parameter redundancy

Inappropriate model structure
```

When convergence problems occur:

- Adjust starting values
- Apply bounds
- Try simpler models
- Compare alternative equations
- Consider biologically meaningful parameter ranges

---

# Biological Reality Matters

The statistically best model is not always
the biologically most meaningful model.

Researchers should consider:

- Biological plausibility
- Parameter interpretation
- Stability of estimates
- Reproducibility
- Experimental context

alongside fit statistics.

---

# Practical Recommendations

## Use Simple Models When

- Sample size is limited
- Fermentation is smooth
- Interpretation is important

Examples:

- Brody
- EXP0
- Ørskov and McDonald

---

## Use Lag Models When

- Colonization delay is expected

Examples:

- EXPL
- Logistic
- Gompertz
- Mitscherlich

---

## Use Flexible Sigmoidal Models When

- Fermentation profiles are complex
- Greater flexibility is desired
- Multiple candidate models are being compared

Examples:

- Groot
- Generalized Michaelis-Menten
- Burr XII
- Inverse Paralogistic
- LE0
- LEL

---

## Use Multi-Pool Models When

- Rapid and slow fractions are biologically relevant
- Substrate heterogeneity is important

Example:

- Dual Logistic

---

# Summary

A successful analysis combines:

- Good model fit
- Biological plausibility
- Parameter interpretability
- Robust convergence

rumenGP provides both classical and
modern approaches to gas-production
modeling, including:

- Brody
- Ørskov and McDonald
- EXP0
- EXPL
- Gompertz
- Logistic
- Mitscherlich
- LE0
- LEL
- Groot
- Generalized Michaelis-Menten
- Dual Logistic
- Burr XII
- Inverse Paralogistic

In addition, the Log-logistic formulation
is already represented mathematically through
the existing Groot and generalized
Michaelis-Menten parameterizations.

Researchers are encouraged to fit multiple
models and evaluate both statistical and
biological performance before selecting
a final representation of fermentation
kinetics.
